Cancer Applications of Choropleth Maps, and the Potential of Cartograms and Alternative Map Displays

### 2020-01-16 review_article.Rmd
figure images

Cancer atlases communicate cancer statistics over geographic domains, typically with a choropleth map. They subdivide these domains into administrative regions such as countries, states, or suburbs. When communicating human-related statistics, the choropleth has a disadvantage in that it draws attention to sparsely populated rural areas to the neglect of small inner city areas. The smaller geographic areas are important to consider if they are densely populated. Alternative map displays, such as a cartogram or a hexagon tile map, can shift the attention of map users from the large rural areas by decreasing their size on the map display. This means alternative displays can be more effective at accurately communicating spatial patterns across spatial areas. It is recommended that alternative displays are included in cancer atlases. In addition, with the ease of today's technology, user interaction with the displays is encouraged. Users should also be able to interactively display different statistics, such as incidence rate or relative incidence, or filtered by demographic variables.

@@

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = "Figure 1: A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atllases are described in Table 1.", fig.align='center', fig.width=12, fig.height=10}
knitr::include_graphics("figures/choropleth_grid.png")
```

@@

```{r ggcartograms, fig.cap = "Figure 2: Common alternatives to maps, showing the same information for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexagon tilemap (non-contiguous). Maps (a) - (c) are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help ease losing information about physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in (d) the hexagon tile map.", message=FALSE, warning=FALSE}
knitr::include_graphics("figures/usa_grid.png")
```

@@

```{r ggtilemap, fig.cap = "Figure 3: Two alternative displays, tile map (left) and geofaceted map (right), showing state age-adjusted rate of incidence for lung and bronchus in the USA. In the tile map, the layout approximates spatial location, with each state being an equal box filled with color representing cancer incidence. The geo-faceted map shows bar charts laid out in a grid approximating the spatial location of the state. The maps show age-adjusted rates for males and females. This display allows the presentation of multiple variables for each geographic area.", out.width="100%"}
knitr::include_graphics("figures/gggrids.png")
```

@@

```{r auscartograms, fig.cap = "Figure 4: Cartograms showing melanoma incidence in Australia: (a) contiguous, partially population transformed, (b) non-contiguous shape preserved, (c) Dorling, (d) hexagon tile map. The contiguous cartogram has expanded the highly populated areas while preserving the full shapes of rural areas. If it accurately sized areas by population, the country would be unrecognizable. The shape-preserved is unreadable due to the small area sizes. The Dorling cartogram presents all areas but many are difficult to compare. The hexagon tile map provides a reasonable spatial distribution despite having isolated hexagons in the outback areas.", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/auscartograms.png")
```

@@

<!-- ## Under-utilized techniques-->

```{r interacting, fig.cap = "Figure 5: Interactive controls of displays in publicly available choropleth cancer maps: (i) GUI controls for statistic, sex, age groups, continents, and cancer types for [Globocan 2018: Cancer Today ](http://globocan.iarc.fr/Pages/Map.aspx), (ii) Menus for variable selection and zooming on [Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/), (iii) Menus for choosing variables and countries in [The Cancer Atlas ](http://canceratlas.cancer.org/data/#?view=map&metric=INCID_ALL_M),  (iv) Tabs for different indicators and cancer types in [Global Cancer Map](http://globalcancermap.com/), (v) Menus and toggles for variable and subset selection in [United States Cancer Statistics: Data Visualizations](https://gis.cdc.gov/Cancer/USCS/DataViz.html).", fig.width = 12, fig.height = 9, results = "asis"}
knitr::include_graphics("figures/interacting.png")
```

@@

```{r animating, fig.cap = "Two examples of advanced interactivity (and animation) in publicly available choropleth cancer maps: a. Linked maps and time-series line plots, with temporal animation in [Map of Cancer Mortality Rates in Spain](http://www.geeitema.org/AtlasET/atlas.jsp?causa=e27MEPOC), b. A highly responsive magnifying glass on a map of [Breast Cancer Mortality in Canada](http://www.ehatlas.ca/light-pollution/maps/breast-cancer-mortality).", fig.width = 12, fig.height = 7, results = "asis"}
knitr::include_graphics("figures/animating.png")
```

>>>>

Cancer atlases communicate cancer statistics over geographic domains, typically with a choropleth map. They subdivide these domains into administrative regions such as countries, states, or suburbs. When communicating human-related statistics, the choropleth has a disadvantage in that it draws attention to sparsely populated rural areas to the neglect of small inner city areas. The smaller geographic areas are important to consider if they are densely populated. Alternative map displays, such as a cartogram or a hexagon tile map, can shift the attention of map users from the large rural areas by decreasing their size on the map display. This means alternative displays can be more effective at accurately communicating spatial patterns across spatial areas. It is recommended that alternative displays are included in cancer atlases. In addition, with the ease of today’s technology, user interaction with the displays is encouraged. Users should also be able to interactively display different statistics, such as incidence rate or relative incidence, or filtered by demographic variables.

@@

```{r choropleth-grid-create, message=FALSE, warning = FALSE, fig.cap = " A selection of choropleth cancer maps from online atlases that are publicly available. Maps of various countries were chosen: United Kingdom, Australia, Spain, USA, Canada, and display several different colour palettes and legends. These atllases are described in Table 1.", fig.align='center', fig.width=12, fig.height=10}
knitr::include_graphics("figures/choropleth_grid.png")
```

@@

```{r ggcartograms, fig.cap = " Common alternatives to maps, showing the same information for the United States of America: (a) contiguous cartogram, (b) non-contiguous, shape-preserved cartogram, (c)  Dorling cartogram (non-contiguous), (d) hexagon tilemap (non-contiguous). Maps (a) - (c) are created by resizing and reshaping the states of the USA to match the 2015 population of the state. This provides a better sense of the extent of disease relative to the population in the country and can help ease losing information about physically small but population-dense states. Map creators give each state equal size and thus equal emphasis in (d) the hexagon tile map.", message=FALSE, warning=FALSE}
knitr::include_graphics("figures/usa_grid.png")
```

@@

```{r ggtilemap, fig.cap = " Two alternative displays, tile map (left) and geofaceted map (right), showing state age-adjusted rate of incidence for lung and bronchus in the USA. In the tile map, the layout approximates spatial location, with each state being an equal box filled with color representing cancer incidence. The geo-faceted map shows bar charts laid out in a grid approximating the spatial location of the state. The maps show age-adjusted rates for males and females. This display allows the presentation of multiple variables for each geographic area.", out.width="100%"}
knitr::include_graphics("figures/gggrids.png")
```

@@

```{r auscartograms, fig.cap = " Cartograms showing melanoma incidence in Australia: (a) contiguous, partially population transformed, (b) non-contiguous shape preserved, (c) Dorling, (d) hexagon tile map. The contiguous cartogram has expanded the highly populated areas while preserving the full shapes of rural areas. If it accurately sized areas by population, the country would be unrecognizable. The shape-preserved is unreadable due to the small area sizes. The Dorling cartogram presents all areas but many are difficult to compare. The hexagon tile map provides a reasonable spatial distribution despite having isolated hexagons in the outback areas.", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/auscartograms.png")
```

@@

```{r interacting, fig.cap = " Interactive controls of displays in publicly available choropleth cancer maps: (i) GUI controls for statistic, sex, age groups, continents, and cancer types for Globocan 2018 [@Globocan], (ii) Menus for variable selection and zooming on Bowel Cancer Australia Atlas, (iii) Menus for choosing variables and countries in The Cancer Atlas, (iv) Tabs for different indicators and cancer types in Global Cancer Map, (v) Menus and toggles for variable and subset selection in United States Cancer Statistics: Data Visualizations.", results = "asis", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/interacting.png")
```

@@

```{r animating, fig.cap = "Two examples of advanced interactivity (and animation) in publicly available choropleth cancer maps: a. Linked maps and time-series line plots, with temporal animation in Map of Cancer Mortality Rates in Spain, b. A highly responsive magnifying glass on a map of Breast Cancer Mortality in Canada.", results = "asis", message=FALSE, warning=FALSE, cache=FALSE, fig.width=8, fig.height=6, dpi=300}
knitr::include_graphics("figures/animating.png")
```
